### 2023-09-30 rec06/img/01.tex
set theory

\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1,xscale=1]
%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300

%Shape: Ellipse [id:dp525262874729056] 
\draw   (312,149.5) .. controls (312,114.43) and (360.58,86) .. (420.5,86) .. controls (480.42,86) and (529,114.43) .. (529,149.5) .. controls (529,184.57) and (480.42,213) .. (420.5,213) .. controls (360.58,213) and (312,184.57) .. (312,149.5) -- cycle ;
%Shape: Circle [id:dp5030750793881615] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (357,159.5) .. controls (357,157.01) and (359.01,155) .. (361.5,155) .. controls (363.99,155) and (366,157.01) .. (366,159.5) .. controls (366,161.99) and (363.99,164) .. (361.5,164) .. controls (359.01,164) and (357,161.99) .. (357,159.5) -- cycle ;
%Shape: Circle [id:dp8729877752431809] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (396,126.5) .. controls (396,124.01) and (398.01,122) .. (400.5,122) .. controls (402.99,122) and (405,124.01) .. (405,126.5) .. controls (405,128.99) and (402.99,131) .. (400.5,131) .. controls (398.01,131) and (396,128.99) .. (396,126.5) -- cycle ;
%Shape: Circle [id:dp7427574956440781] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (447,176.5) .. controls (447,174.01) and (449.01,172) .. (451.5,172) .. controls (453.99,172) and (456,174.01) .. (456,176.5) .. controls (456,178.99) and (453.99,181) .. (451.5,181) .. controls (449.01,181) and (447,178.99) .. (447,176.5) -- cycle ;
%Shape: Circle [id:dp6662281997079922] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (452,122.5) .. controls (452,120.01) and (454.01,118) .. (456.5,118) .. controls (458.99,118) and (461,120.01) .. (461,122.5) .. controls (461,124.99) and (458.99,127) .. (456.5,127) .. controls (454.01,127) and (452,124.99) .. (452,122.5) -- cycle ;
%Shape: Circle [id:dp7807995256741351] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (399,181.5) .. controls (399,179.01) and (401.01,177) .. (403.5,177) .. controls (405.99,177) and (408,179.01) .. (408,181.5) .. controls (408,183.99) and (405.99,186) .. (403.5,186) .. controls (401.01,186) and (399,183.99) .. (399,181.5) -- cycle ;
%Shape: Circle [id:dp656685899722891] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (426,153.5) .. controls (426,151.01) and (428.01,149) .. (430.5,149) .. controls (432.99,149) and (435,151.01) .. (435,153.5) .. controls (435,155.99) and (432.99,158) .. (430.5,158) .. controls (428.01,158) and (426,155.99) .. (426,153.5) -- cycle ;
%Shape: Circle [id:dp8189169790009678] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (490,150.5) .. controls (490,148.01) and (492.01,146) .. (494.5,146) .. controls (496.99,146) and (499,148.01) .. (499,150.5) .. controls (499,152.99) and (496.99,155) .. (494.5,155) .. controls (492.01,155) and (490,152.99) .. (490,150.5) -- cycle ;
%Shape: Circle [id:dp520965594846118] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (349,119.5) .. controls (349,117.01) and (351.01,115) .. (353.5,115) .. controls (355.99,115) and (358,117.01) .. (358,119.5) .. controls (358,121.99) and (355.99,124) .. (353.5,124) .. controls (351.01,124) and (349,121.99) .. (349,119.5) -- cycle ;
%Shape: Circle [id:dp854731305701526] 
\draw  [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ,fill opacity=1 ] (227,153.5) .. controls (227,151.01) and (229.01,149) .. (231.5,149) .. controls (233.99,149) and (236,151.01) .. (236,153.5) .. controls (236,155.99) and (233.99,158) .. (231.5,158) .. controls (229.01,158) and (227,155.99) .. (227,153.5) -- cycle ;

% Text Node
\draw (388,103) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle x$};
% Text Node
\draw (213,127) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle x$};
% Text Node
\draw (267,143) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \in $};

\end{tikzpicture}

### 2023-09-30 rec06/paper.tex
set theory

\documentclass{amsart}
\input{decls}
\title{Sets}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Set Theory}
\label{sec:basic-set-theory}

Set theory made its debut in Cantor's 1874 paper: ``On a Property of the Collection of All Real Algebraic Numbers.''
The Cantorian set theory contained several contradictions such as Russell's paradox, but mathematicians did not abandon set theory entirely.
The work of Zermelo, Fraenkel, and Skolem resulted in a well-known axiomatizations of set theory.

In this class, we will not talk about axiomatic set theory.
Instead, we will look at set theory na\"ively.

\subsection{Set}
\label{sec:set}

The atomic objects in set theory are sets.
A set can be thought of as a collection of dots.
These dots have no internal structures, so the only sensible question that one can ask is ``is $x$ a member of $y$?''

\begin{center}
  \input{img/01}
\end{center}

The language of set theory contains a binary predicate symbol $\in$.
This is known as the \emph{membership relation}.
$x \in y$ means $x$ is an element of $y$.

\subsection{Equality}
\label{sec:equality}

\subsection{Subsets}
\label{sec:subsets}

\subsection{Comprehension}
\label{sec:comprehension}

\subsection{Union}
\label{sec:union}

\subsection{Intersections}
\label{sec:intersections}

\subsection{von Neumann Ordinals}
\label{sec:von-neumann-ordinals}

\end{document}

### 2023-09-30 rec06/paper.tex
comprehension

@@
 
-\begin{center}
-  \input{img/01}
-\end{center}
-
 The language of set theory contains a binary predicate symbol $\in$.
@@
 
+\subsection{Subsets}
+\label{sec:subsets}
+
+\begin{defn}
+  A set $x$ is a subset of $y$, denoted $x \subseteq y$, if every element in $x$ is also in $y$.
+  \[
+    x \subseteq y \cng \forall z.(z \in x \imp z \in y)
+  \]
+  The relation $\subseteq$ is called \emph{set inclusion}.
+\end{defn}
+
 \subsection{Equality}
@@
 
-\subsection{Subsets}
-\label{sec:subsets}
+Two sets are equal when they contain the same elements.
+We can express this in terms of the set inclusion relation.
+\[
+  \forall x. \forall y. ((x \subseteq y \wedge y \subseteq x) \imp x = y)
+\]
+Given two sets $x$ and $y$, to prove that $x = y$, it suffices to prove $x \subseteq y$ and $y \subseteq x$.
 
@@
 
+Given a set $w$, there is a \emph{subcollection} of $w$ whose elements satisfy certain property $\varphi$.
+\[
+  \{ x \in w \mid \varphi(x) \}
+\]
+We declare that such subcollections are sets.
+For example, given the set of all natural numbers $\dN$,
+\[
+  \{ x \in \dN \mid \mathrm{even}(x) \}
+\]
+is a subcollection of all even natural numbers.
+
 \subsection{Union}
@@
 
-\subsection{Intersections}
-\label{sec:intersections}
+\subsection{Intersection}
+\label{sec:intersection}
 

### 2023-10-01 rec06/paper.tex
subsets

@@
 Set theory made its debut in Cantor's 1874 paper: ``On a Property of the Collection of All Real Algebraic Numbers.''
-The Cantorian set theory contained several contradictions such as Russell's paradox, but mathematicians did not abandon set theory entirely.
-The work of Zermelo, Fraenkel, and Skolem resulted in a well-known axiomatizations of set theory.
+Later, mathematicians such as Russell realized that the Cantorian set theory contained several contradictions, but mathematicians did not abandon set theory.
+The work of Zermelo, Fraenkel, and Skolem resulted in a well-known axiomatization of set theory.
 
@@
 
-The atomic objects in set theory are sets.
-A set can be thought of as a collection of dots.
-These dots have no internal structures, so the only sensible question that one can ask is ``is $x$ a member of $y$?''
-
-The language of set theory contains a binary predicate symbol $\in$.
+Intuitively, a set is a collection of elements.
+These elements have no internal structures, so you can think of a set as a collection of dots, and the only question that one can ask is ``is $x$ a member of $y$?''
+In other words, the language of set theory contains a binary predicate symbol $\in$.
 This is known as the \emph{membership relation}.
 $x \in y$ means $x$ is an element of $y$.
+
+\begin{eg}
+  \begin{enumerate}
+  \item[]
+  \item The empty set: $\varnothing$.
+  \item The set containing the empty set: $\{\varnothing\}$.
+  \item A set containing 3 elements: $\{ a,b,c \}$.
+  \item The set of all natural numbers: $\dN = \{ 0,1,2,\ldots \}$.
+  \item The set of all integers: $\dZ = \{ \ldots,-2,-1,0,1,2,\ldots \}$.
+  \item The set containing $\dN$ and $\dZ$: $\{ \dN,\dZ \}$.
+  \end{enumerate}
+\end{eg}
+
+\begin{eg}
+  \begin{enumerate}
+  \item[]
+  \item Nothing is in the empty set: $x \notin \varnothing$.
+  \item The set containing the empty set has an element: $\varnothing \in \{\varnothing\}$.
+  \item $a \in \{ a, b, c \}$, $b \in \{ a, b, c \}$, $c \in \{ a, b, c \}$.
+  \item $\dN \in \{ \dN, \dZ \}$, $\dZ \in \{ \dN, \dZ \}$, but $0 \notin \{ \dN,\dZ \}$.
+  \end{enumerate}
+\end{eg}
 
@@
 \end{defn}
+
+\begin{eg}
+  The empty set is a subset of any set.
+\end{eg}
+\begin{proof}
+  Let $y$ be any set.
+  By definition, $\varnothing \subseteq y \cng \forall z.(z \in \varnothing \imp z \in y)$.
+  Let $z$ be given.
+  Assume that $z \in \varnothing$.
+  This allows us to derive a contradiction since $z \notin \varnothing$.\footnote{This is not a proof by contradiction, but let's ignore the details for now.}
+\end{proof}
 
@@
 \]
-is a subcollection of all even natural numbers.
+is the subset of all even natural numbers.
 

### 2023-10-01 rec06/paper.tex
comprehension

@@
 
-\begin{eg}
+\begin{lem}
   The empty set is a subset of any set.
-\end{eg}
+  \[
+    \forall y.\,\varnothing \subseteq y
+  \]
+\end{lem}
 \begin{proof}
@@
   Let $z$ be given.
-  Assume that $z \in \varnothing$.
-  This allows us to derive a contradiction since $z \notin \varnothing$.\footnote{This is not a proof by contradiction, but let's ignore the details for now.}
+  Assume that $z \in \varnothing$, but this is impossible since $z \notin \varnothing$.\footnote{This is not a proof by contradiction, but let's ignore the details for now.}
+\end{proof}
+
+\begin{lem}
+  Every set is a subset of itself.
+\end{lem}
+\begin{proof}
+  Exercise.
 \end{proof}
@@
 Given two sets $x$ and $y$, to prove that $x = y$, it suffices to prove $x \subseteq y$ and $y \subseteq x$.
+
+\begin{eg}
+  \begin{enumerate}
+  \item[]
+  \item $\{a,b,c,d,d\} = \{a,b,c,d\}$.
+  \item $\{a,b,c\} = \{c,b,a\}$.
+  \end{enumerate}
+\end{eg}
 
@@
 
+\subsection{Set Difference}
+\label{sec:set-difference}
+
 \subsection{von Neumann Ordinals}

### 2023-10-02 rec06/paper.tex
power set

@@
 These elements have no internal structures, so you can think of a set as a collection of dots, and the only question that one can ask is ``is $x$ a member of $y$?''
-In other words, the language of set theory contains a binary predicate symbol $\in$.
+
+The language of set theory contains a binary predicate symbol $\in$.
 This is known as the \emph{membership relation}.
@@
   Every set is a subset of itself.
+  \[
+    \forall x.\, x \subseteq x
+  \]
 \end{lem}
@@
 \end{proof}
+
+\begin{eg}\label{eg:subsets}
+  \begin{enumerate}
+  \item[]
+  \item $\varnothing$ has a subset $\varnothing$.
+  \item $\{ \varnothing \}$ has subsets $\varnothing$ and $\{ \varnothing \}$.
+  \item $\{ a,b \}$ has subsets $\varnothing$, $\{ a \}$, $\{ b \}$, and $\{ a, b \}$.
+  \item $\{ a,b,c \}$ has subsets $\varnothing$, $\{ a \}$, $\{ b \}$, $\{ c \}$, $\{ a,b \}$, $\{ a,c \}$, $\{ b,c \}$, and $\{ a,b,c \}$.
+  \end{enumerate}
+\end{eg}
 
@@
 We declare that such subcollections are sets.
-For example, given the set of all natural numbers $\dN$,
+
+\begin{eg}
+  \begin{enumerate}
+  \item[]
+  \item The set of all even natural numbers: $\{ x \in \dN \mid \mathrm{even}(x) \}$.
+  \item The set of all odd natural numbers: $\{ x \in \dN \mid \mathrm{odd}(x) \}$.
+  \item The set of all integers divisible by 2: $\{ x \in \dZ \mid x \cng 0 \mod 2 \}$.
+  \item The set of all real numbers between $0$ and $1$ (inclusive): $\{ x \in \dR \mid 0 \leq x \leq 1 \}$.
+  \end{enumerate}
+\end{eg}
+
+\subsection{Power Set}
+\label{sec:power-set}
+
+In \cref{eg:subsets}, $\{ a,b,c \}$ has subsets $\varnothing$, $\{ a \}$, $\{ b \}$, $\{ c \}$, $\{ a,b \}$, $\{ a,c \}$, $\{ b,c \}$, and $\{ a,b,c \}$.
+These subsets form a collection
 \[
-  \{ x \in \dN \mid \mathrm{even}(x) \}
+  \{ \varnothing, \{ a \}, \{ b \}, \{ c \}, \{ a,b \}, \{ a,c \}, \{ b,c \}, \{ a,b,c \} \}
 \]
-is the subset of all even natural numbers.
+We declare that such a collection is a set.
+
+\begin{defn}
+  Let $x$ be a set.
+  The \emph{power set} of $x$, denoted $\cP(x)$, is the set of all subsets of $x$.
+\end{defn}
+
+\begin{eg}\label{eg:subsets}
+  \begin{enumerate}
+  \item[]
+  \item $\cP(\varnothing) = \{\varnothing\}$.
+  \item $\cP(\{ \varnothing \}) = \{\varnothing, \{\varnothing\}\}$.
+  \item $\cP(\{\varnothing, \{\varnothing\}\}) = \{ \varnothing, \{\varnothing\}, \{\{\varnothing\}\}, \{\varnothing, \{\varnothing\}\} \}$.
+  \item $\cP(\{ a,b,c \}) = \{ \varnothing, \{ a \}, \{ b \}, \{ c \}, \{ a,b \}, \{ a,c \}, \{ b,c \}, \{ a,b,c \} \}$.
+  \end{enumerate}
+\end{eg}
+
+\begin{thm}[Cantor's Theorem]
+  For any set $x$, there is no surjective function $f : x \to \cP(x)$.
+\end{thm}
 
@@
 
-\subsection{von Neumann Ordinals}
-\label{sec:von-neumann-ordinals}
-
 \end{document}

### 2023-10-03 rec06/paper.tex
intersection

@@
 
-Given a set $w$, there is a \emph{subcollection} of $w$ whose elements satisfy certain property $\varphi$.
+Given a set $w$, there is a set of $w$ whose elements satisfy certain property $\varphi$.
 \[
@@
 \]
-We declare that such subcollections are sets.
 
@@
 In \cref{eg:subsets}, $\{ a,b,c \}$ has subsets $\varnothing$, $\{ a \}$, $\{ b \}$, $\{ c \}$, $\{ a,b \}$, $\{ a,c \}$, $\{ b,c \}$, and $\{ a,b,c \}$.
-These subsets form a collection
+These subsets form a set
 \[
@@
 \]
-We declare that such a collection is a set.
 
@@
 
+\begin{defn}
+  Given a nonempty set of sets $\cF$, the \emph{union}, denoted $\cup\cF$, is a set whose elements are those elements of \emph{some} set in $\cF$.
+\end{defn}
+
+\begin{eg}
+  \begin{enumerate}
+  \item[]
+  \item If $\cF = \{\{1,2,3\},\{a,b,c\}\}$ then $\cup\cF = \{1,2,3,a,b,c\}$.
+  \item If $\cF = \{\{a,b,c\},\{b,c,d\},\{c,d,e\}\}$ then $\cup\cF = \{a,b,c,d,e\}$.
+  \end{enumerate}
+\end{eg}
+
+\begin{notn}
+  When $\cF$ has just a few elements, for example $\cF = \{\{1,2,3\},\{a,b,c\}\}$, we write
+  \[
+    \{1,2,3\} \cup \{a,b,c\}
+  \]
+  for $\cup\cF$.
+\end{notn}
+
 \subsection{Intersection}
 \label{sec:intersection}
+
+\begin{defn}
+  Given a nonempty set of sets $\cF$, the \emph{intersection}, denoted $\cap\cF$, is a set whose elements are those elements of \emph{every} set in $\cF$.
+\end{defn}
+
+\begin{eg}
+  \begin{enumerate}
+  \item[]
+  \item If $\cF = \{\{1,2,3\},\{a,b,c\}\}$ then $\cap\cF = \varnothing$.
+  \item If $\cF = \{\{a,b,c\},\{b,c,d\},\{c,d,e\}\}$ then $\cap\cF = \{c\}$.
+  \end{enumerate}
+\end{eg}
+
+\begin{notn}
+  When $\cF$ has just a few elements, for example $\cF = \{\{1,2,3\},\{a,b,c\}\}$, we write
+  \[
+    \{1,2,3\} \cap \{a,b,c\}
+  \]
+  for $\cap\cF$.
+\end{notn}
 

### 2023-10-04 rec-function/paper.tex
functions

\documentclass{amsart}
\input{decls}
\title{Functions}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Countable Sets and Uncountable Sets}
\label{sec:countable-sets-and-uncountable-sets}

\begin{thm}\label{thm:function-space-uncountable}
  $\dN^{\dN}$ is uncountable.
\end{thm}
\begin{proof}
  Suppose that $\dN^{\dN}$ is countable, i.e., $\dN \iso \dN^{\dN}$.
  One possible interpretations of this hypothesis is that every function $f : \dN \to \dN$ can be given a unique natural-number code.
  That is, there are functions
  \begin{align}
    \mathsf{decode} &: \dN \to \dN^{\dN}\\
    \mathsf{encode} &: \dN^{\dN} \to \dN
  \end{align}
  that are mutual inverses.
  Consider the function
  \begin{align}
    k &: \dN \to \dN\\
    k &: n \mapsto \mathsf{decode}(n)(n) + 1
  \end{align}
  Given a code $n$, the function $k$ decodes $n$, yielding a function $\dN \to \dN$, then evaluates that function at $n$, and finally adds 1 to the result.

  The function $k$ has a unique code given by $\mathsf{encode}(k)$.
  Now, let's evaluate $k$ at its own code:
  \begin{align}
    k(\mathsf{encode}(k)) &= \mathsf{decode}(\mathsf{encode}(k))(\mathsf{encode}(k)) + 1\\
                          &= k(\mathsf{encode}(k)) + 1
  \end{align}
  This is a contradiction.\footnote{This proof is not a proof by contradiction. In fact, it is a constructively valid proof.}
\end{proof}

\cref{thm:function-space-uncountable} tells us that some functions $f : \dN \to \dN$ are uncomputable: there are only countably many programs that one can write, but there are uncountably many endofunctions on $\dN$. Thus, some of those functions do not have a corresponding program that computes it.

\end{document}

### 2023-10-15 rec08/paper.tex
relations

--- 2023-10-04 rec-function/paper.tex
@@
 \input{decls}
-\title{Functions}
+\title{Functions and Relations}
 \author{Frank Tsai}
@@
 \tableofcontents
+
+\section{Relations}
+\label{sec:relations}
+
+\begin{defn}
+  An $n$-ary \emph{relation} $R$ on a set $S$ is a subset:
+  \[
+    R \subseteq S^{n}
+  \]
+  We write $R(a,\ldots,z)$ whenever $(a,\ldots,z) \in R$.
+
+  Binary relations will be the main focus of this class.
+  For these relations, it is customary to use infix notations.
+  That is, we write $aRb$ instead of $R(a,b)$.
+\end{defn}
+
+\begin{eg}
+  The less-than-or-equal-to relation $\leq$ on $\dN$ is the subset
+  \[
+    \{(0,0), (0,1), \ldots, (1,1), (1,2), \ldots\} \subseteq \dN \times \dN
+  \]
+\end{eg}
+
+\begin{eg}
+  The divisibility relation $\mid$ on $\dZ$ is defined by
+  \[
+    a \mid b \iff \exists c.\,b = ac
+  \]
+  It is the subset
+  \[
+    \{(a,b) \in \dZ \times \dZ \mid \exists c.\,b = ac\}
+  \]
+\end{eg}
+
+\begin{eg}
+  The adjacency relation in a simple graph: two vertices $u$ are $v$ are adjacent if they are connected by an edge.
+  It is the subset
+  \[
+    \{(u,v) \in V \times V \mid (u, v) \in E \vee (v, u) \in E\}
+  \]
+\end{eg}
+
+\begin{defn}[Reflexivity]
+  A binary relation $R$ on a set $S$ is \emph{reflexive} if for all elements $a$ of $S$, $aRa$.
+  \[
+    \forall a.\,aRa
+  \]
+\end{defn}
+
+\begin{defn}[Symmetry]
+  A binary relation $R$ on a set $S$ is \emph{symmetric} if for any two elements $a,b$ of $S$, if $aRb$ then $bRa$.
+  \[
+    \forall a.\,\forall b.\,aRb \imp bRa
+  \]
+\end{defn}
 
@@
   Suppose that $\dN^{\dN}$ is countable, i.e., $\dN \iso \dN^{\dN}$.
-  One possible interpretations of this hypothesis is that every function $f : \dN \to \dN$ can be given a unique natural-number code.
+  A possible interpretation of this hypothesis is that every function $f : \dN \to \dN$ can be given a unique natural-number code.
   That is, there are functions
@@
   \end{align}
-  This is a contradiction.\footnote{This proof is not a proof by contradiction. In fact, it is a constructively valid proof.}
+  This is a contradiction.
 \end{proof}

### 2023-10-15 rec08/paper.tex
relations

@@
 \begin{defn}
-  An $n$-ary \emph{relation} $R$ on a set $S$ is a subset:
+  An $n$-ary \emph{relation} $R$ on a set $S$ can be encoded as a subset:
   \[
@@
   We write $R(a,\ldots,z)$ whenever $(a,\ldots,z) \in R$.
-
   Binary relations will be the main focus of this class.
@@
 \begin{eg}
-  The less-than-or-equal-to relation $\leq$ on $\dN$ is the subset
+  The substring relation $\sqsubseteq$ on $\{a,b\}^{*}$ is the subset
   \[
-    \{(0,0), (0,1), \ldots, (1,1), (1,2), \ldots\} \subseteq \dN \times \dN
+    \{(\varepsilon,\varepsilon), (\varepsilon, a), \ldots, (a, a), (a, ab), (a, ba), \ldots\}
   \]
@@
 \begin{eg}
-  The adjacency relation in a simple graph: two vertices $u$ are $v$ are adjacent if they are connected by an edge.
+  The adjacency relation on a simple graph: two vertices $u$ are $v$ are adjacent if they are connected by an edge.
   It is the subset
@@
 \begin{defn}[Reflexivity]
-  A binary relation $R$ on a set $S$ is \emph{reflexive} if for all elements $a$ of $S$, $aRa$.
+  A binary relation $R$ on a set $S$ is \emph{reflexive} if every element of $S$ is related to itself by $R$.
   \[
@@
 
+\begin{eg}
+  The divisibility relation on $\dZ$ is reflexive because every integer divides into itself once.
+\end{eg}
+
 \begin{defn}[Symmetry]
-  A binary relation $R$ on a set $S$ is \emph{symmetric} if for any two elements $a,b$ of $S$, if $aRb$ then $bRa$.
+  A binary relation $R$ on a set $S$ is \emph{symmetric} if whenever $a$ is related to $b$ by $R$, then $b$ is also related to $a$ by $R$.
   \[
-    \forall a.\,\forall b.\,aRb \imp bRa
+    \forall a.\forall b.\,(aRb \imp bRa)
   \]
 \end{defn}
+
+\begin{eg}
+  The adjacency relation on a simple graph is symmetric.
+  If a vertex $u$ is adjacent to another vertex $v$, then $v$ is also adjacent to $u$.
+\end{eg}
+
+\begin{defn}[Transitivity]
+  A binary relation $R$ on a set $S$ is \emph{transitive} if for any three elements $a,b,c$ of $S$, if $aRb$ and $bRc$ then $aRc$.
+  \[
+    \forall a.\forall b.\forall c.\,(aRb \wedge bRc \imp aRc)
+  \]
+\end{defn}
+
+\begin{eg}
+  The substring relation on $\{a,b\}^{*}$ is transitive.
+  In fact, it is also reflexive, but it is not symmetric.
+\end{eg}
+
+\begin{defn}[Equivalence Relation]
+  A binary relation $R$ on a set $S$ is an \emph{equivalence relation} if it is
+  \begin{enumerate}
+  \item reflexive,
+  \item symmetric, and
+  \item transitive.
+  \end{enumerate}
+\end{defn}
+
+\begin{prop}\label{prop:cng-mod-equiv}
+  The congruence-modulo-2 relation on $\dZ$ is defined by
+  \[
+    a \cng b \mod{2} \iff 2 \mid (a - b)
+  \]
+  It is an equivalence relation.
+\end{prop}
+\begin{proof}
+  (Reflexivity). Let $a$ be any integer.
+  We need to prove that $a \cng a \mod{2}$.
+  By definition, this is equivalent to proving $2 \mid (a - a)$, or equivalently, $2 \mid 0$.
+  By definition again, this is equivalent to $\exists c.\,0 = 2c$.
+  Setting $c := 0$ yields $0 = 2 \cdot 0 = 0$ as desired.
+
+  (Symmetry). Let $a, b$ be any integers.
+  Assume that $a \cng b \mod{2}$.
+  By definition, this hypothesis asserts that there's an integer $c$ so that $a - b = 2c$.
+  We need to prove $\exists k.\,b - a = 2k$.
+  Setting $k := -c$ yields $b - a = -(a - b) = -2c = 2(-c)$ as desired.
+
+  (Transitivity). Let $a, b, c$ be any integers.
+  Assume that $a \cng b \mod{2}$ and that $b \cng c \mod{2}$.
+  By definition, these two hypotheses assert that there are integers $n, m$ so that $a - b = 2n$ and $c - b = 2m$.
+  We need to show that $\exists c.\,a - c = 2c$.
+  Setting $c := n - m$ yields $2(n - m) = 2n - 2m = a - b - (c - b) = a - b - c + b = a - c$ as desired.
+\end{proof}
+
+\begin{defn}[Antisymmetry]
+  A binary relation $R$ on a set $S$ is \emph{antisymmetric} if for any two elements $a, b$ of $S$, if $aRb$ and $bRa$ then $a = b$.
+\end{defn}
+
+\begin{eg}
+  The subset relation $\subseteq$ on $\cP(S)$ is antisymmetric.
+  Recall that two sets $A$ and $B$ are equal precisely when $A \subseteq B$ and $B \subseteq A$.
+\end{eg}
+
+\begin{rmk}
+  Antisymmetry does \textbf{not} imply \textbf{a}symmetry.
+  For example, the indiscrete relation $I$ on the singleton set $\{a\}$, defined as
+  \[
+    I = \{(a,a)\}
+  \]
+  is both antisymmetric and symmetric.
+\end{rmk}
+
+\begin{defn}[Preorder]
+  A binary relation is a \emph{preorder} if it is
+  \begin{enumerate}
+  \item reflexive, and
+  \item transitive.
+  \end{enumerate}
+\end{defn}
+
+\begin{defn}[Partial Order]
+  A \emph{partial order} is a preorder that additionally satisfies antisymmetry.
+\end{defn}
+
+\begin{prop}\label{prop:div-n-partial}
+  The divisibility relation on $\dN$ is a partial order.
+\end{prop}
+\begin{proof}
+  (Reflexivity): Exercise.
+
+  (Transitivity): Exercise.
+  Hint: See \cref{prop:cng-mod-equiv}.
+
+  (Antisymmetry): Let $a, b$ be natural numbers so that $a \mid b$ and $b \mid a$.
+  These hypotheses assert that there are natural numbers $n, m$ so that $b = an$ and that $a = bm$.
+  Thus, $b = (bm)n$.
+  If $b = 0$, then since $a = bm = 0m = 0$, $a = b$ as desired.
+  However, if $b \ne 0$, then $mn = 1$.
+  Since $n, m$ are natural numbers, $n = m = 1$.
+  Thus, $a = b$ as desired.
+\end{proof}
+
+\begin{rmk}
+  \cref{prop:div-n-partial} does not hold if we replace $\dN$ with $\dZ$ because $2 \mid -2$ and $-2 \mid 2$, but $2 \ne -2$.
+  Although the divisibility relation on $\dZ$ is not a partial order, it is a preorder.
+\end{rmk}
+
+\section{Functions}
+\label{sec:functions}
 

### 2023-10-21 rec08/paper.tex
fixed error
fixed name error
functions

@@
   Assume that $a \cng b \mod{2}$ and that $b \cng c \mod{2}$.
-  By definition, these two hypotheses assert that there are integers $n, m$ so that $a - b = 2n$ and $c - b = 2m$.
-  We need to show that $\exists c.\,a - c = 2c$.
-  Setting $c := n - m$ yields $2(n - m) = 2n - 2m = a - b - (c - b) = a - b - c + b = a - c$ as desired.
+  By definition, these two hypotheses assert that there are integers $n, m$ so that $a - b = 2n$ and $b - c = 2m$.
+  We need to show that $\exists k.\,a - c = 2k$.
+  Setting $k := n + m$ yields $2(n + m) = 2n + 2m = (a - b) + (b - c) = a - b + b - c = a - c$ as desired.
 \end{proof}
@@
 
+Intuitively, a function from a set $A$ to another set $B$ is a rule for assigning each element of $A$ a unique element of $B$.
+
 \section{Countable Sets and Uncountable Sets}

### 2023-10-21 rec08/paper.tex
cantor thm

@@
 
-Intuitively, a function from a set $A$ to another set $B$ is a rule for assigning each element of $A$ a unique element of $B$.
+Intuitively, a function is a rule for assigning each element of a set to a unique element of another set.
+In set theory, functions can be encoded as a special binary relation.
+
+\begin{defn}
+  A binary relation $R \subseteq A \times B$ is (left) \emph{total} if
+  \[
+    \forall a \in A.\exists b \in B.\,(a,b) \in R
+  \]
+\end{defn}
+
+\begin{defn}
+  A binary relation $R \subseteq A \times B$ is \emph{functional} if
+  \[
+    \forall a \in A.\forall b \in B.\forall c \in C.\,((a,b) \in R \wedge (a,c) \in R \imp b = c)
+  \]
+\end{defn}
+
+\begin{defn}
+  A function $f$ from a set $A$ to another set $B$, denoted $f : A \to B$ is a binary relation
+  \[
+    f \subseteq A \times B
+  \]
+  such that $f$ is \emph{total} and \emph{functional}.
+  We write $f(a) = b$ for $(a,b) \in f$.
+  Writing the two conditions in this notation is perhaps more illuminating:
+  \begin{enumerate}
+  \item Totality:
+    \[
+      \forall a \in A.\exists b \in B.\,f(a) = b
+    \]
+  \item Functionality:
+    \[
+      \forall a \in A.\forall b \in B.\forall c \in C.\,(f(a) = b \wedge f(a) = c \imp b = c)
+    \]
+  \end{enumerate}
+  The set $A$ is called the \emph{domain} of $f$, and the set $B$ is called the \emph{codomain} of $f$.
+\end{defn}
+
+\begin{thm}[Functional Extensionality]
+  Two functions $f,g : A \to B$ are equal if and only if $f(a) = g(a)$ for all $a \in A$.
+\end{thm}
+\begin{proof}
+  The ``only if'' direction is obvious.
+  For the ``if'' direction, assume that $f(a) = g(a)$ for all $a \in A$.
+  To prove that $f = g$, it suffices to prove $f \subseteq g$ and $g \subseteq f$.
+  Now, suppose that $(a,b) \in f$.
+  Since $f(a) = g(a)$, $(a, g(a)) \in f$.
+  By functionality, $g(a) = b$.
+  Thus, $(a,b) \in g$, proving that $f \subseteq g$.
+  The proof of $g \subseteq f$ is completely analogous.
+\end{proof}
+
+\begin{defn}
+  Given two functions $f : A \to B$ and $g : B \to C$, the composition $g \circ f : A \to C$ (reads ``$g$ after $f$'') is a function defined by
+  \[
+    (g \circ f)(x) = g(f(x))
+  \]
+  Note that $g \circ f$ is defined only if the codomain of $f$ and the domain of $g$ are the same.
+\end{defn}
+
+\begin{lem}
+  Composition is associative, i.e., $(f \circ g) \circ h = f \circ (g \circ h)$.
+\end{lem}
+\begin{proof}
+  Exercise.
+  Hint: Use functional extensionality.
+\end{proof}
+
+\begin{defn}
+  For any set $S$, there is a special function $\id_{S}$, called the \emph{identity function on $S$}, defined by
+  \[
+    \id_{S}(s) = s
+  \]
+\end{defn}
+
+\begin{lem}
+  For any function $f : A \to B$, $\id_{B} \circ f = f$ and $f \circ \id_{A} = f$.
+\end{lem}
+\begin{proof}
+  Exercise.
+\end{proof}
+
+\begin{defn}
+  A function $f : A \to B$ is \emph{injective}, denoted $f : A \mono B$, if
+  \[
+    \forall a \in A.\forall a' \in A.\,(f(a) = f(a') \imp a = a')
+  \]
+\end{defn}
+
+\begin{defn}
+  A function $f : A \to B$ is \emph{surjective}, denoted $f : A \epi B$, if
+  \[
+    \forall b \in B.\exists a \in A.\,f(a) = b
+  \]
+\end{defn}
+
+\begin{thm}[Cantor's Theorem]
+  For any set $S$, there is no surjective functions $f : S \epi \cP(S)$.
+\end{thm}
+\begin{proof}
+  Suppose that $f : S \epi \cP(S)$.
+  Consider the subset $\{s \in S \mid s \notin f(s)\}$.
+  Since $f$ is surjective, there must be some $s' \in S$ so that $f(s') = \{s \in S \mid s \notin f(s)\}$.
+  If $s' \in f(s')$, then by definition, $s' \notin f(s')$.
+  Similarly, if $s' \notin f(s')$, then by definition, $s' \in f(s')$.
+  This is a contradiction.
+\end{proof}
+
+\begin{defn}
+  A function $f : A \to B$ is \emph{bijective} if it is injective and surjective.
+\end{defn}
 
@@
 \label{sec:countable-sets-and-uncountable-sets}
+
+\begin{defn}
+  A set $S$ is \emph{countable} if there is a bijection $f : S \to \dN$.
+\end{defn}
 

### 2023-10-22 rec08/paper.tex
functions

@@
 
-\begin{lem}
+\begin{lem}\label{lem:associativity}
   Composition is associative, i.e., $(f \circ g) \circ h = f \circ (g \circ h)$.
@@
 
-\begin{lem}
+\begin{lem}\label{lem:unitality}
   For any function $f : A \to B$, $\id_{B} \circ f = f$ and $f \circ \id_{A} = f$.
@@
 \end{proof}
+
+\cref{lem:associativity,lem:unitality} mean that sets and functions between them assemble into a category.
+Category theory is an interesting subject that we will sadly not discuss in this class.
 
@@
 \end{defn}
+
+\begin{defn}
+  A function $f : A \to B$ is \emph{invertible} if there is a function $g : B \to A$ such that
+  \begin{enumerate}
+  \item $f \circ g = \id_{B}$, and
+  \item $g \circ f = \id_{A}$.
+  \end{enumerate}
+  $g$ is called the inverse of $f$.
+  When $f$ is invertible, we write $f\inv$ for its inverse.
+\end{defn}
+
+\begin{thm}
+  A function $f : A \to B$ is invertible if and only if $f$ is bijective.
+\end{thm}
+\begin{proof}
+  The ``if'' direction: assume that $f$ is invertible.
+  Then there is a function $f\inv : B \to A$ such that $f \circ f\inv = \id_{B}$ and $f\inv \circ f = \id_{A}$.
+
+  (Injectivity): Let $a, a' \in A$ be given.
+  Assume that $f(a) = f(a')$.
+  Then $\id_{A}(a) = f\inv(f(a)) = f\inv(f(a')) = \id_{A}(a')$.
+  Thus, $a = a'$.
+
+  (Surjectivity): Let $b \in B$ be given.
+  We need to show that there is some $a \in A$ so that $f(a) = b$.
+  Choose $a := f\inv(b)$, then $f(f\inv(b)) = \id_{B}(b) = b$.
+
+  The ``only if'' direction: assume that $f$ is bijective.
+  We need to show that $f$ is invertible.
+  To this end, we construct a function $f\inv : B \to A$.
+  For each $a \in A$ so that $f(a) = b$, we take $f\inv(b) = a$.
+  To show that $f\inv$ is a function, we must show that it is total and functional.
+  Totality follows from surjectivity of $f$ and functionality follows from injectivity of $f$.
+  The details are left to the reader as an exercise.
+\end{proof}
 

### 2023-10-28 induction/paper.lagda.tex
removed junk

\documentclass{amsart}
\usepackage{agda}
\usepackage{fancyvrb}
\input{decls}
\title{Induction}
\author{Frank Tsai}
\date{\today}
% \thanks{}

\DefineVerbatimEnvironment
{code}{Verbatim}
{}

\begin{document}
\maketitle
\tableofcontents

\section{Induction}
\label{sec:induction}

Roughly speaking, natural numbers $\dN$ are the ``minimal'' data type equipped with a successor function $s : \dN \to \dN$.
We can define $\dN$ inductively:
\begin{enumerate}
\item $0$ is a natural number;
\item if $n$ is a natural number, then $s(n)$ is also a natural number.
\end{enumerate}

Consider the following thought experiment:
If I know $P(0)$ and $P(0) \imp P(1)$, then I can conclude $P(1)$.
\begin{mathpar}
  \inferrule
  { P(0) \\\\ P(0) \imp P(1) }
  { P(1) }
\end{mathpar}
If I additionally know $P(1) \imp P(2)$, then I can also conclude $P(2)$.
\begin{mathpar}
  \inferrule
  { P(0) \\\\ P(0) \imp P(1) \\\\ P(1) \imp P(2) }
  { P(2) }
\end{mathpar}
Now suppose that I know $P(0)$ and $P(0) \imp P(1), P(1) \imp P(2), \ldots$.
In other words, I have the following hypotheses:
\begin{mathpar}
  P(0) \and \forall k.\,(P(k) \imp P(s(k)))
\end{mathpar}
Now, if I want to prove $\forall n.\,P(n)$, I need to introduce a variable $n$ and prove $P(n)$.
Since $n$ is a natural number, it has to be either $0$ or $s(m)$ for some natural number $m$.
If $n$ is $0$, then proving $P(0)$ is easy because it follows directly from the first hypothesis.
If $n$ is $s(m)$, then the second hypothesis says it suffices to prove $P(m)$.
To prove this, we can recursively apply the argument just described, i.e., do a case analysis on $m$.
Each time, the input to $P$ become smaller.

You may have noticed that this is just a recursive function with a base case.
In fact, we can implement this in a programming language called Agda.

\end{document}

### 2023-10-28 induction/induction.tex
induction

\documentclass{amsart}
\input{decls}
\usepackage{agda}
\title{Induction}
\author{Frank Tsai}
\date{\today}
% \thanks{}

\usepackage{newunicodechar}
\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
\newunicodechar{←}{\ensuremath{\mathnormal\from}}
\newunicodechar{→}{\ensuremath{\mathnormal\to}}
\newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
\newunicodechar{≡}{\ensuremath{\mathnormal\cng}}
\newunicodechar{≐}{\ensuremath{\mathnormal=}}
\newunicodechar{ℕ}{\ensuremath{\mathnormal\dN}}

\begin{document}
\maketitle
\tableofcontents

\section{Induction}
\label{sec:induction}

Roughly speaking, natural numbers $\dN$ are the ``minimal'' data type equipped with a successor function $s : \dN \to \dN$.
We can define $\dN$ inductively:
\begin{enumerate}
\item $0$ is a natural number;
\item if $n$ is a natural number, then $s(n)$ is also a natural number.
\end{enumerate}

\begin{code}%
\>[0]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaPrimitive{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[0][@{}l@{\AgdaIndent{0}}]%
\>[2]\AgdaInductiveConstructor{zero}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\<%
\\
%
\>[2]\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{ℕ}\<%
\end{code}

We may want to prove that every natural number has some property $P$.
For example, we may want to prove that for any natural number $n$, $2 \mid n(n+1)$.
If we proceed directly, we may write down ``let $n$ be a natural number, we need to prove $2 \mid n(n+1)$.''
Sadly, we are stuck because we don't know what $n$ is.

The good news is we know that every natural number is either $0$ or the successor of some other natural number, i.e., $s(m)$.
If we can somehow show the followings:
\begin{mathpar}
  P(0) \and \forall k.\,(P(k) \imp P(s(k)))
\end{mathpar}
then we can reason as follows: let $n$ be any natural number, we want to show $P(n)$.
We can do a case analysis on $n$.
If $n$ is $0$, then we need to show $P(0)$, which we have shown already.
If $n$ is $s(m)$ instead, then we need to show $P(s(m))$.
Since we know $\forall k.\,(P(k) \imp P(s(k)))$, it suffices to show $P(m)$.
To show $P(m)$, we repeat the previous argument, i.e., we do a case analysis on $m$ and so on.

This is similar to writing a recursive function with a base case in programming.
In fact, we can implement this in programming languages such as Agda.

\begin{code}%
%
\>[2]\AgdaFunction{induction}\AgdaSpace{}%
\AgdaSymbol{:}%
\>[23I]\AgdaSymbol{\{}\AgdaBound{P}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaPrimitive{Set}\AgdaSymbol{\}}\AgdaSpace{}%
\AgdaSymbol{→}\<%
\\
\>[.][@{}l@{}]\<[23I]%
\>[14]\AgdaBound{P}\AgdaSpace{}%
\AgdaNumber{0}\AgdaSpace{}%
\AgdaSymbol{→}\<%
\\
%
\>[14]\AgdaSymbol{(∀}\AgdaSpace{}%
\AgdaBound{k}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}%
\AgdaBound{k}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaBound{k}\AgdaSymbol{))}\AgdaSpace{}%
\AgdaSymbol{→}\<%
\\
%
\>[14]\AgdaSymbol{∀}\AgdaSpace{}%
\AgdaBound{n}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}%
\AgdaBound{n}\<%
\\
%
\>[2]\AgdaFunction{induction}\AgdaSpace{}%
\AgdaBound{b}\AgdaSpace{}%
\AgdaBound{f}\AgdaSpace{}%
\AgdaInductiveConstructor{zero}\AgdaSpace{}%
\AgdaSymbol{=}\AgdaSpace{}%
\AgdaBound{b}\<%
\\
%
\>[2]\AgdaFunction{induction}\AgdaSpace{}%
\AgdaBound{b}\AgdaSpace{}%
\AgdaBound{f}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaBound{x}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{=}\AgdaSpace{}%
\AgdaBound{f}\AgdaSpace{}%
\AgdaBound{x}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaFunction{induction}\AgdaSpace{}%
\AgdaBound{b}\AgdaSpace{}%
\AgdaBound{f}\AgdaSpace{}%
\AgdaBound{x}\AgdaSymbol{)}\<%
\end{code}

In summary, the principle of induction says that to prove that $\forall n.\,P(n)$, it suffices to prove two things:
\begin{enumerate}
\item Base case: $P(0)$, and
\item Induction step: $\forall k.\,(P(k) \imp P(s(k)))$.
\end{enumerate}
To prove the induction step, we introduce a natural number $k$ and assume $P(k)$ (the induction hypothesis), and derive $P(s(k))$.

\section{Examples}
\label{sec:examples}

\begin{prop}
  For all $n \in \dN$, $2 \mid n(n + 1)$.
\end{prop}
\begin{proof}
  $P(n)$ is $2 \mid n(n + 1)$.
  By induction on $n$, it suffices to prove
  \begin{enumerate}
  \item Base case: $P(0)$, i.e., $2 \mid 0(0 + 1)$.
  \item Induction step: $\forall k.\,(P(k) \imp P(k + 1))$, i.e.,
    \[ \forall k.\,(2 \mid k(k + 1) \imp 2 \mid (k + 1)((k + 1) + 1)) \]
  \end{enumerate}

  (Base case): Clearly, $2$ divides $0$.

  (Induction step): Let $k$ be a natural number.
  Assume $2 \mid k(k + 1)$.
  We need to show that $2 \mid (k + 1)((k + 1) + 1)$, or equivalently, $2 \mid (k(k+1) + 2k + 2)$.
  By the induction hypothesis, $2$ divides $k(k + 1)$, and clearly, $2$ also divides $2k + 2$.
\end{proof}

\begin{prop}
  For all $n \in \dN$, $2^{0} + 2^{1} + \cdots + 2^{n} = 2^{n + 1} - 1$, i.e.,
  \[
    \sum_{i=0}^{n}2^{i} = 2^{n + 1} - 1
  \]
\end{prop}
\begin{proof}
  $P(n)$ is $2^{0} + 2^{1} + \cdots + 2^{n} = 2^{n + 1} - 1$.
  By induction on $n$, it suffices to prove
  \begin{enumerate}
  \item Base case: $P(0)$, i.e., $2^{0} = 2^{0 + 1} - 1$.
  \item Induction step: $\forall k.\,(P(k) \imp P(k + 1))$, i.e.,
    \[ \forall k.\,\left(\sum_{i=0}^{n}2^{i} = 2^{n + 1} - 1 \imp \sum_{i=0}^{n+1}2^{i} = 2^{(n+1) + 1} - 1\right) \]
  \end{enumerate}

  (Base case): It follows immediately by computation.

  (Induction case): Let $k$ be a natural number.
  Assume that
  \[
    \sum_{i=0}^{n}2^{i} = 2^{n + 1} -1
  \]
  We need to prove
  \[
    \sum_{i=0}^{n+1}2^{i} = 2^{(n+1) + 1} - 1
  \]
  Using the induction hypothesis, the left hand side can be rewritten as follows:
  \[
    \left(\sum_{i=0}^{n}2^{i}\right) + 2^{n+1} = 2^{n + 1} - 1 + 2^{n + 1} = 2^{n + 2} - 1
  \]
\end{proof}

So far, the choice of $P$ is straightforward.
Let's see an example where the choice of $P$ is not so straightforward, but let's see what happens if we choose the na\"ive $P$.

\begin{prop}\label{prop:recurrence}
  Consider a function $f : \dN \to \dN$ defined recursively as follows:
  \begin{align}
    f(0) &= 1\\
    f(1) &= 3\\
    f(n+2) &= 2f(n + 1) - f(n)
  \end{align}
  This function has a closed form:
  \[
    \forall n.\,f(n) = 2n + 1
  \]
\end{prop}
\begin{proof}[Failed Attempt]
  $P(n)$ is $f(n) = 2n + 1$.
  By induction on $n$, we need to prove the base case and the induction step.

  (Base case): $P(0)$ is $f(0) = 2 \cdot 0 + 1$.
  By definition, $f(0) = 1$ and by computation $2 \cdot 0 + 1 = 1$, so the base case goes through fine.

  (Induction step): Let $k \in \dN$.
  Assume that $f(k) = 2k + 1$, we need to prove $f(k+1) = 2(k+1) + 1$.
  If $k$ is $0$, then the equality follows by computation.
  If $k > 0$, then by definition $f(k + 1) = 2f(k) - f(k - 1)$.
  By the induction hypothesis, $f(k) = 2k + 1$, so
  \[
    f(k + 1) = 2(2k + 1) - f(k - 1)
  \]
  We are stuck because the induction hypothesis does not tell us anything about $f(k - 1)$.
\end{proof}

We need a stronger induction hypothesis that tells us something about $f(k - 1)$.
This requires a different choice of $P$.
Let's consider the following lemma.

\begin{lem}\label{lem:stronger-IH}
  For all $m \in \dN$ and $n \in \dN$, if $n < m$ then $f(n) = 2n + 1$.
\end{lem}
\begin{proof}
  $P(m)$ is $\forall n.\,(n < m \imp f(n) = 2n + 1)$.
  By induction on $m$, it suffices to show the base case and the induction step.

  (Base case): $P(0)$ is $\forall n.\,(n < 0 \imp f(n) = 2n + 1)$.
  Let $n \in \dN$.
  Assume $n < 0$.
  This is a contradiction because no natural number is strictly less than $0$.

  (Induction step): Let $k \in \dN$.
  The induction hypothesis $P(k)$ is
  \[
    \forall n.\,(n < k \imp f(n) = 2n + 1)
  \]
  and we need to prove $P(k+1)$, which is
  \[
    \forall n.\,(n < k + 1 \imp f(n) = 2n + 1)
  \]
  Note that the induction hypothesis now tells us something about $f(n)$ for any $n$ less than $k$.
  To prove $P(k+1)$, let $n \in \dN$.
  Assume $n < k + 1$.

  If $k$ is $0$ or $1$, then $n$ is $0$ or $1$.
  These two cases follow directly from how $f$ is defined.

  For $k > 1$, there are two cases:
  If $n < k$, then the result follows immediately from the induction hypothesis.

  If $n = k$, then by definition $f(k) = 2f(k-1) - f(k-2)$.\footnote{Note that $k-2$ and $k-1$ are natural numbers because $k > 1$ in this case. We eliminated $k = 0$ and $k = 1$ in the previous step for this step to work.}
  Since $k - 2 < k$ and $k - 1 < k$, the induction hypothesis says that $f(k-2) = 2(k-2) + 1 = 2k - 3$ and that $f(k-1) = 2(k-1) + 1 = 2k - 1$.
  Thus,
  \begin{align}
    f(k) &= 2f(k-1) - f(k-2)\\
         &= 2(2k - 1) - (2k - 3)\\
         &= 4k - 2 - 2k + 3\\
         &= 2k + 1
  \end{align}
\end{proof}

\cref{prop:recurrence} is an immediate corollary of \cref{lem:stronger-IH}.
\begin{proof}[Proof of \cref{prop:recurrence}]
  Let $n \in \dN$.
  We need to prove $f(n) = 2n + 1$.
  This follows immediately from \cref{lem:stronger-IH} by setting $m := n + 1$.
\end{proof}

You may have heard \emph{strong induction} in class.
Unfortunately, the name ``strong induction'' is somewhat misleading because anything provable with strong induction can be proved with mathematical induction presented here and vice-versa, i.e., strong induction is \textbf{not} stronger than mathematical induction.
In fact, the pattern used in \cref{lem:stronger-IH} is what strong induction does.

\end{document}

### 2023-10-29 induction/induction.tex
strong induction

@@
 
-  If $n = k$, then by definition $f(k) = 2f(k-1) - f(k-2)$.\footnote{Note that $k-2$ and $k-1$ are natural numbers because $k > 1$ in this case.
-We eliminated $k = 0$ and $k = 1$ in the previous step for this step to work.}
+  If $n = k$, then by definition $f(k) = 2f(k-1) - f(k-2)$.
   Since $k - 2 < k$ and $k - 1 < k$, the induction hypothesis says that $f(k-2) = 2(k-2) + 1 = 2k - 3$ and that $f(k-1) = 2(k-1) + 1 = 2k - 1$.
@@
 
+\begin{prop}\label{prop:lincomb}
+  For any $n \in \dN$, if $n \geq 2$ then $n$ is a linear combination of $2$ and $3$, i.e., there are natural numbers $i$ and $j$ so that $n = 2i + 3j$.
+\end{prop}
+
+Again, we need to strengthen the induction hypothesis.
+
+\begin{lem}\label{lem:lincomb-strong}
+  For any $m \in \dN$ and $n \in \dN$, if $n < m$ then if additionally $n \geq 2$ then $n$ is a linear combination of $2$ and $3$.
+\end{lem}
+\begin{proof}
+  $P(m)$ is $\forall n.\,(n < m \imp (n \geq 2 \imp \exists i.\,\exists j.\,n = 2i + 3j))$.
+
+  (Base case): $P(0)$ is $\forall n.\,(n < 0 \imp (n \geq 2 \imp \exists i.\,\exists j.\,n = 2i + 3j))$.
+  Let $n \in \dN$.
+  Assume $n < 0$.
+  This is a contradiction.
+
+  (Induction step): Let $k \in \dN$.
+  Assume $P(k)$, i.e.,
+  \[
+    \forall n.\,(n < k \imp (n \geq 2 \imp \exists i.\,\exists j.\,n = 2i + 3j))
+  \]
+  We need to prove $P(k+1)$, i.e.,
+  \[
+    \forall n.\,(n < k + 1 \imp (n \geq 2 \imp \exists i.\,\exists j.\,n = 2i + 3j))
+  \]
+  Let $n \in \dN$.
+  Assume $n < k + 1$ and $n \geq 2$, so $k \geq 2$.
+  If $k = 2$ then $n$ has to be $2$, which can be expressed as $2 \cdot 1 + 3 \cdot 0$.
+  If $k = 3$ then $n$ has to be $2$ or $3$.
+  We know how to express $2$, and $3$ can be expressed as $2 \cdot 0 + 3 \cdot 1$.
+  For $k \geq 4$, if $n < k$ then the induction hypothesis gives us what we want.
+  If $n = k$, then consider $k - 2$.
+  The induction hypothesis tells us that there are natural numbers $a$ and $b$ so that $k - 2 = 2a + 3b$.
+  We can then express $n$ as $n = k = k - 2 + 2 = 2(a + 1) + 3b$.
+\end{proof}
+
+\begin{proof}[Proof of \cref{prop:lincomb}]
+  Let $n \in \dN$.
+  Assume $n \geq 2$.
+  The result follows immediately from \cref{lem:lincomb-strong}.
+\end{proof}
+
 \end{document}

### 2023-11-10 graphs/paper.tex
added graphs

\documentclass{amsart}
\input{decls}
\title{Graphs}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Graphs}
\label{sec:graphs}

\begin{defn}
  \label{defn:graphs}
  A (directed) \emph{graph} $G$ consists of the following data:
  \begin{enumerate}
  \item A set $V$ of \emph{vertices}.
  \item A set $E \subseteq V \times V$ of \emph{edges}.
  \end{enumerate}
\end{defn}

\begin{rmk}
  \label{rmk:edges-are-binary-relations}
  Recall that a binary relation on a set $V$ can be encoded as any subset of $V \times V$.
  Thus, a graph is a set $V$ equipped with a binary relation $E$.
\end{rmk}

\begin{eg}
  \label{eg:graphs-example1}
  Let $V = \{A, B, C\}$ and $E = \{(A,B),(B,A),(B,C),(C,A)\}$.
  \begin{center}
    \begin{tikzpicture}[node distance= 2cm]
      \begin{scope}[every node/.style={circle,thick,draw}]
        \node (A) {$A$};
        \node (B) [below left of= A] {$B$};
        \node (C) [below right of=A] {$C$};
      \end{scope}
      \begin{scope}[every edge/.style={draw= black, very thick}]
        \path[->] (A) edge[bend right= 30] (B);
        \path[->] (B) edge[bend right= 30] (A);
        \path[->] (B) edge (C);
        \path[->] (C) edge (A);
      \end{scope}
    \end{tikzpicture}
  \end{center}
\end{eg}

\begin{defn}
  \label{defn:graphs-undirected}
  A graph $(V,E)$ is \emph{undirected} if the binary relation $E$ is symmetric.
\end{defn}

\begin{rmk}
  \label{rmk:example1-not-undirected}
  \cref{eg:graphs-example1} is not an undirected graph since $(B,C) \in E$, but $(C,B) \notin E$.
  Similarly, $(C,A) \in E$, but $(A,C) \notin E$.
\end{rmk}

\begin{rmk}
  \label{rmk:drop-arrows-in-undirected-graphs}
  In an undirected graph, we drop the arrow tips as they convey no additional information.
\end{rmk}

\begin{eg}
  \label{eg:graphs-example2}
  Let $V = \{A, B, C\}$ and $E = V \times V$.
  Note that $E$ is symmetric.
  \begin{center}
    \begin{tikzpicture}[node distance= 2cm]
      \begin{scope}[every node/.style={circle,thick,draw}]
        \node (A) {$A$};
        \node (B) [below left of= A] {$B$};
        \node (C) [below right of=A] {$C$};
      \end{scope}
      \begin{scope}[every edge/.style={draw= black, very thick}]
        \path (A) edge (B);
        \path (A) edge (C);
        \path (B) edge (C);
      \end{scope}
    \end{tikzpicture}
  \end{center}
\end{eg}

\begin{defn}
  Let $G$ be a graph.
  A \emph{walk} is a sequence of vertices and edges defined inductively as follows:
  \begin{enumerate}
  \item If $(v_{1},v_{2}) \in E$ then the sequence $v_{1},(v_{1},v_{2}),v_{2}$ is a walk.
  \item If $(v_{n}, v_{m}) \in E$ and $v_{1},(v_{1},v_{2}),v_{2},\ldots,v_{n}$ is a walk then $v_{1},(v_{1},v_{2}),v_{2},\ldots,v_{n},(v_{n},v_{m}),v_{m}$ is a walk.
  \end{enumerate}
\end{defn}

\begin{eg}
  \label{eg:walk-example1}
  Let $G$ be the graph defined in \cref{eg:graphs-example2}.
  The sequence
  \[
    A,(A,B),B,(B,C),C,(C,A),A,(A,B),B
  \]
  is a walk.
  \begin{center}
    \begin{tikzpicture}[node distance= 2cm]
      \begin{scope}[every node/.style={circle,thick,draw}]
        \node (A) {$A$};
        \node (B) [below left of= A] {$B$};
        \node (C) [below right of=A] {$C$};
      \end{scope}
      \begin{scope}
        [every edge/.style={draw= black, very thick},
        every node/.style={fill=white,circle}]
        \path (A) edge (B);
        \path (A) edge (C);
        \path (B) edge (C);
        \path[->] (A) edge[draw= red, bend right= 30] node {$1$} (B);
        \path[->] (B) edge[draw= red, bend right= 30] node {$2$} (C);
        \path[->] (C) edge[draw= red, bend right= 30] node {$3$} (A);
        \path[->] (A) edge[draw= red, bend left= 30] node {$4$} (B);
      \end{scope}
    \end{tikzpicture}
  \end{center}
  Each red arrow represents a step of the walk.
\end{eg}

\section{BFS}
\label{sec:bfs}

\section{DFS}
\label{sec:dfs}

\end{document}

### 2023-11-11 graphs/paper.tex
bfs

@@
 \input{decls}
+\usetikzlibrary {arrows.meta,backgrounds,fit,positioning}
 \title{Graphs}
@@
 \begin{defn}
-  Let $G$ be a graph.
+  \label{defn:walks}
+  Let $G = (V,E)$ be a graph.
   A \emph{walk} is a sequence of vertices and edges defined inductively as follows:
   \begin{enumerate}
+  \item The empty sequence is a walk.
   \item If $(v_{1},v_{2}) \in E$ then the sequence $v_{1},(v_{1},v_{2}),v_{2}$ is a walk.
-  \item If $(v_{n}, v_{m}) \in E$ and $v_{1},(v_{1},v_{2}),v_{2},\ldots,v_{n}$ is a walk then $v_{1},(v_{1},v_{2}),v_{2},\ldots,v_{n},(v_{n},v_{m}),v_{m}$ is a walk.
+  \item If $(v_{n}, v_{m}) \in E$ and $v_{1},(v_{1},v_{2}),v_{2},\ldots,v_{n}$ is a walk then the sequence
+    \[
+      v_{1},(v_{1},v_{2}),v_{2},\ldots,v_{n},(v_{n},v_{m}),v_{m}
+    \]
+    is a walk.
   \end{enumerate}
@@
 
+\begin{defn}
+  \label{defn:walks-length}
+  The \emph{length} of a walk is the number of edges in that walk.
+\end{defn}
+
 \begin{eg}
-  \label{eg:walk-example1}
+  \label{eg:walk-example-1}
   Let $G$ be the graph defined in \cref{eg:graphs-example2}.
@@
 
-\section{BFS}
-\label{sec:bfs}
+\begin{defn}
+  \label{defn:paths}
+  Let $G = (V,E)$ be a graph.
+  A \emph{path} is a walk in which all vertices are distinct.
+\end{defn}
 
-\section{DFS}
-\label{sec:dfs}
+\begin{ceg}
+  \label{ceg:nonpath-example-1}
+  The walk defined in \cref{eg:walk-example-1} is not a path since the vertices $A$ and $B$ are repeated.
+\end{ceg}
+
+\begin{eg}
+  \label{eg:path-example-1}
+  Let $G$ be the graph defined in \cref{eg:graphs-example2}.
+  The sequence
+  \[
+    A,(A,B),B,(B,C),C
+  \]
+  is a path.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [below left of= A] {$B$};
+        \node (C) [below right of=A] {$C$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (C);
+        \path[->] (A) edge[draw= red, bend right= 30] node {$1$} (B);
+        \path[->] (B) edge[draw= red, bend right= 30] node {$2$} (C);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+\end{eg}
+
+\begin{defn}
+  \label{defn:reachability}
+  Let $G = (V,E)$ be a graph.
+  For any two vertices $u,v \in V$, $u$ and $v$ are said to be \emph{s-t reachable} if there is a path from $u$ to $v$.
+\end{defn}
+
+\begin{eg}
+  \label{eg:reachability-example-1}
+  Consider the following (undirected) graph:
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [below of= A] {$B$};
+        \node (C) [right of= B] {$C$};
+        \node (D) [right of= A] {$D$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (C);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+  $A$ and $C$ are s-t reachable.
+  There are two paths connecting them (blue and red, respectively):
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [below of= A] {$B$};
+        \node (C) [right of= B] {$C$};
+        \node (D) [right of= A] {$D$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[draw= red] (B);
+        \path (A) edge[draw= blue] (C);
+        \path (B) edge[draw= red] (C);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+\end{eg}
+
+\begin{ceg}
+  \label{ceg:reachability-nonexample-1}
+  Consider the same graph defined in \cref{eg:reachability-example-1}, the vertices $A$ and $D$ are not s-t reachable.
+\end{ceg}
+
+\begin{thm}
+  \label{thm:reachability-is-an-equivalence}
+  Let $G = (V, E)$ be an undirected graph.
+  The s-t reachability relation defined in \cref{defn:reachability} is an equivalence relation on the set of vertices $V$.
+\end{thm}
+\begin{proof}
+  (Reflexivity): Let $v \in V$.
+  The empty walk is trivially a path.
+  Thus, $v$ is s-t reachable from itself.
+
+  (Symmetry): Let $u,v \in V$ be any two vertices.
+  Assume that $v$ is s-t reachable from $u$, i.e., there is a path $p$ from $u$ to $v$.
+  We need to prove that $u$ is s-t reachable from $v$, i.e., there is a path from $v$ to $u$.
+  To this end, we do induction on the length of $p$.
+  In the base case, $p$ has length 0, so it is the empty path.
+  Thus, the result follows from reflexivity.
+  In the induction step, $p$ is a path of length $k+1$ and it is of the form $p',(u',v),v$, where $u'$ is some vertex and $p'$ is a path from $u$ to $u'$ of length $k$.
+  By the induction hypothesis, there is a path $\widetilde{p}$ from $u'$ to $u$.
+  Since $G$ is undirected, $(v,u')$ is also an edge of $G$.
+  Thus, the sequence
+  \[
+    v,(v,u'),\widetilde{p}
+  \]
+  defines a path from $v$ to $u$.
+
+  (Transitivity): Let $u,v,w \in V$ be any three vertices.
+  Assume that there is a path $p$ from $u$ to $v$ and a path $p'$ from $v$ to $w$.
+  The concatenation of $p$ and $p'$ defines a path from $u$ to $w$.
+  You can prove this rigorously by induction on the length of $p$.
+  The detail is left as an exercise.
+\end{proof}
+
+\begin{rmk}
+  \label{rmk:reachability-not-equivalence-for-directed-graphs}
+  In general, \cref{thm:reachability-is-an-equivalence} fails to hold for directed graphs because the s-t reachability relation is, in general, not symmetric.
+\end{rmk}
+
+\begin{ceg}
+  \label{ceg:reachability-not-equivalence-for-directed-graphs-nonexample-1}
+  Consider the following graph.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [right of= A] {$B$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path[->] (A) edge (B);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+  There is a path from $A$ to $B$, but there is no path from $B$ to $A$.
+\end{ceg}
+
+\begin{defn}
+  \label{defn:connected-components}
+  Let $G$ be an undirected graph.
+  A \emph{(path) connected component} of $G$ is an equivalence class with respect to the s-t reachability relation.
+\end{defn}
+
+\begin{eg}
+  \label{eg:connected-components-example-1}
+  The graph defined in \cref{eg:reachability-example-1} has exactly 2 connected components (red and blue, respectively).
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [below of= A] {$B$};
+        \node (C) [right of= B] {$C$};
+        \node (D) [right of= A] {$D$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (C);
+      \end{scope}
+      \begin{scope}[on background layer]
+        \node[circle, draw=blue, very thick, fill= blue!30, fit= (D)] {};
+      \end{scope}
+      \begin{scope}[on background layer]
+        \filldraw[red!30, draw=red, very thick] (-0.5,-2.5) -- (2.5,-2.5) -- (2.7,-2) -- (0.5,0.5) -- (-0.5,0.5) -- cycle;
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+\end{eg}
+
+\begin{defn}
+  \label{defn:connected-graphs}
+  An undirected graph $G$ is said to be \emph{(path) connected} if it has exactly one connected component.
+\end{defn}
+
+\begin{eg}
+  \label{eg:connected-graphs-example-1}
+  The graph defined in \cref{eg:graphs-example2} is connected because it has exactly one connected component.
+\end{eg}
+
+\begin{ceg}
+  \label{ceg:connected-graphs-nonexample-1}
+  The graph defined in \cref{eg:reachability-example-1} is not connected because it has two connected components.
+\end{ceg}
+
+\section{Graph exploration algorithms}
+\label{sec:graph-exploration-algorithms}
+
+\begin{rmk}
+  \label{rmk:graph-exploration-algorithms-main-idea-1}
+  To explore every vertex of an undirected graph $(V,E)$, we divide $V$ into two piles: frontier and unexplored.
+\end{rmk}
+
+\begin{rmk}
+  \label{rmk:graph-exploration-algorithms-main-idea-2}
+  If the unexplored pile is empty, then that means every vertex has been explored.
+  We can remove everything from the frontier pile.
+\end{rmk}
+
+\begin{rmk}
+  \label{rmk:graph-exploration-algorithms-main-idea-3}
+  If the frontier pile is empty, move an arbitrary vertex in the unexplored pile to the frontier pile.
+\end{rmk}
+
+\begin{rmk}
+  \label{rmk:graph-exploration-algorithms-main-idea-4}
+  If the frontier pile is not empty, pick the next vertex $v$ in the frontier pile and move every vertex adjacent to $v$ in the unexplored pile to the frontier pile.
+  At this point, $v$ has been fully explored, so we can remove $v$ from the frontier pile.
+\end{rmk}
+
+\begin{rmk}
+  \label{rmk:graph-exploration-algorithms-next-vertex}
+  The ``next vertex'' in \cref{rmk:graph-exploration-algorithms-main-idea-4} is intentionally abstract.
+  There are two natural ways to pick the ``next vertex:'' pick the first vertex or the last vertex added to the frontier pile.
+  The former choice corresponds to breadth-first search (BFS) and the latter choice corresponds to depth-first choice (DFS).
+\end{rmk}
+
+\begin{eg}
+  \label{eg:bfs-example}
+  Consider the following graph:
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C) [below right of= A] {$C$};
+        \node (D) [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (C);
+        \path (B) edge (D);
+        \path (B) edge (E);
+        \path (C) edge (D);
+        \path (C) edge (E);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+  We run BFS on this graph.
+
+  (Step 1): The frontier pile is empty and the unexplored pile consists of every vertex in the graph.
+  By \cref{rmk:graph-exploration-algorithms-main-idea-3} we move an arbitrary vertex, e.g, $B$, to the frontier pile.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node[blue] (B) [above right of= A] {$B$};
+        \node (C) [below right of= A] {$C$};
+        \node (D) [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (C);
+        \path (B) edge (D);
+        \path (B) edge (E);
+        \path (C) edge (D);
+        \path (C) edge (E);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  (Step 2): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $B$.
+  We have to move $A$, $C$, $D$, and $E$ to the frontier pile and remove $B$ from the frontier pile.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node[blue] (A) {$A$};
+        \node[gray] (B) [above right of= A] {$B$};
+        \node[blue] (C) [below right of= A] {$C$};
+        \node[blue] (D) [right of= B] {$D$};
+        \node[blue] (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge (C);
+        \path (B) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (B) edge[red] (E);
+        \path (C) edge (D);
+        \path (C) edge (E);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  (Step 3): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $A$.
+  We do not have to move anything to the frontier pile because it is not adjacent to any unexplored vertex.
+  Finally, we need to remove $A$ from the frontier pile.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node[gray] (A) {$A$};
+        \node[gray] (B) [above right of= A] {$B$};
+        \node[blue] (C) [below right of= A] {$C$};
+        \node[blue] (D) [right of= B] {$D$};
+        \node[blue] (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[gray] (C);
+        \path (B) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (B) edge[red] (E);
+        \path (C) edge (D);
+        \path (C) edge (E);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  (Step 4): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $C$.
+  Again, we do not have to move anything to the frontier pile because it is not adjacent to any unexplored vertex.
+  Finally, we need to remove $C$ from the frontier pile.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node[gray] (A) {$A$};
+        \node[gray] (B) [above right of= A] {$B$};
+        \node[gray] (C) [below right of= A] {$C$};
+        \node[blue] (D) [right of= B] {$D$};
+        \node[blue] (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[gray] (C);
+        \path (B) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (B) edge[red] (E);
+        \path (C) edge[gray] (D);
+        \path (C) edge[gray] (E);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  (Step 5): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $D$.
+  We need to move $F$ to the frontier pile and remove $D$ from the frontier pile.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node[gray] (A) {$A$};
+        \node[gray] (B) [above right of= A] {$B$};
+        \node[gray] (C) [below right of= A] {$C$};
+        \node[gray] (D) [right of= B] {$D$};
+        \node[blue] (E) [right of= C] {$E$};
+        \node[blue] (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[gray] (C);
+        \path (B) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (B) edge[red] (E);
+        \path (C) edge[gray] (D);
+        \path (C) edge[gray] (E);
+        \path (D) edge[gray] (E);
+        \path (D) edge[red] (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  (Step 6): The unexplored pile is empty.
+  By \cref{rmk:graph-exploration-algorithms-main-idea-2}, we can remove everything from the frontier pile.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node[gray] (A) {$A$};
+        \node[gray] (B) [above right of= A] {$B$};
+        \node[gray] (C) [below right of= A] {$C$};
+        \node[gray] (D) [right of= B] {$D$};
+        \node[gray] (E) [right of= C] {$E$};
+        \node[gray] (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[gray] (C);
+        \path (B) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (B) edge[red] (E);
+        \path (C) edge[gray] (D);
+        \path (C) edge[gray] (E);
+        \path (D) edge[gray] (E);
+        \path (D) edge[red] (F);
+        \path (E) edge[gray] (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+\end{eg}
 

### 2023-11-11 graphs/paper.tex
graphs

@@
 
-\begin{rmk}
-  \label{rmk:edges-are-binary-relations}
+\para
+\label{para:edges-are-binary-relations}
 Recall that a binary relation on a set $V$ can be encoded as any subset of $V \times V$.
 Thus, a graph is a set $V$ equipped with a binary relation $E$.
-\end{rmk}
 
@@
 
-\begin{rmk}
-  \label{rmk:example1-not-undirected}
+\para
 \cref{eg:graphs-example1} is not an undirected graph since $(B,C) \in E$, but $(C,B) \notin E$.
 Similarly, $(C,A) \in E$, but $(A,C) \notin E$.
-\end{rmk}
 
-\begin{rmk}
-  \label{rmk:drop-arrows-in-undirected-graphs}
+\para
 In an undirected graph, we drop the arrow tips as they convey no additional information.
-\end{rmk}
 
@@
 \end{thm}
-\begin{proof}
-  (Reflexivity): Let $v \in V$.
+\begin{proof}\hfill
+  \para
+  Reflexivity: Let $v \in V$.
   The empty walk is trivially a path.
@@
 
-  (Symmetry): Let $u,v \in V$ be any two vertices.
+  \para
+  Symmetry: Let $u,v \in V$ be any two vertices.
   Assume that $v$ is s-t reachable from $u$, i.e., there is a path $p$ from $u$ to $v$.
@@
 
-  (Transitivity): Let $u,v,w \in V$ be any three vertices.
+  \para
+  Transitivity: Let $u,v,w \in V$ be any three vertices.
   Assume that there is a path $p$ from $u$ to $v$ and a path $p'$ from $v$ to $w$.
@@
 
+\para
+\label{para:graph-exploration-algorithms-main-idea-1}
+To explore every vertex of an undirected graph $(V,E)$, we divide $V$ into three piles: explored, frontier, and unexplored.
+Initially, the unexplored pile contains every vertex and the explored and frontier piles are empty.
+
+\para
+\label{para:graph-exploration-algorithms-main-idea-5}
+We can explore the graph recursively.
+We match the current piles against the conditions stated in
+\ref{para:graph-exploration-algorithms-main-idea-2},
+\ref{para:graph-exploration-algorithms-main-idea-3}, and
+\ref{para:graph-exploration-algorithms-main-idea-4}
+in this order.
+
+\para
+\label{para:graph-exploration-algorithms-main-idea-2}
+If the unexplored pile is empty, we can move every vertex in the frontier pile to the explored pile.
+This is the base case of the recursion.
+
+\para
+\label{para:graph-exploration-algorithms-main-idea-3}
+If the frontier pile is empty, move an arbitrary vertex in the unexplored pile to the frontier pile.
+Make a recursive call on the current piles (i.e., go back to \ref{para:graph-exploration-algorithms-main-idea-5}).
+
+\para
+\label{para:graph-exploration-algorithms-main-idea-4}
+If the frontier pile is not empty, pick the next vertex $v$ in the frontier pile.
+If $v$ is not adjacent to any unexplored vertex, then $v$ has been fully explored, so we can move it to the explore pile.
+Otherwise, we move an arbitrary vertex adjacent to $v$ in the unexplored pile to the frontier pile.
+Finally, we make a recursive call on the current piles (i.e., go back to \ref{para:graph-exploration-algorithms-main-idea-5}).
+
 \begin{rmk}
-  \label{rmk:graph-exploration-algorithms-main-idea-1}
-  To explore every vertex of an undirected graph $(V,E)$, we divide $V$ into two piles: frontier and unexplored.
+  \label{rmk:graph-exploration-algorithms-next-vertex}
+  The ``next vertex'' in \ref{para:graph-exploration-algorithms-main-idea-4} is kept abstract intentionally.
+  There are two natural ways to pick the ``next vertex:'' pick the first vertex or the last vertex added to the frontier pile.
+  The former choice corresponds to breadth-first search (BFS) and the latter choice corresponds to depth-first search (DFS).
 \end{rmk}
@@
 \begin{rmk}
-  \label{rmk:graph-exploration-algorithms-main-idea-2}
-  If the unexplored pile is empty, then that means every vertex has been explored.
-  We can remove everything from the frontier pile.
-\end{rmk}
-
-\begin{rmk}
-  \label{rmk:graph-exploration-algorithms-main-idea-3}
-  If the frontier pile is empty, move an arbitrary vertex in the unexplored pile to the frontier pile.
-\end{rmk}
-
-\begin{rmk}
-  \label{rmk:graph-exploration-algorithms-main-idea-4}
-  If the frontier pile is not empty, pick the next vertex $v$ in the frontier pile and move every vertex adjacent to $v$ in the unexplored pile to the frontier pile.
-  At this point, $v$ has been fully explored, so we can remove $v$ from the frontier pile.
-\end{rmk}
-
-\begin{rmk}
-  \label{rmk:graph-exploration-algorithms-next-vertex}
-  The ``next vertex'' in \cref{rmk:graph-exploration-algorithms-main-idea-4} is intentionally abstract.
-  There are two natural ways to pick the ``next vertex:'' pick the first vertex or the last vertex added to the frontier pile.
-  The former choice corresponds to breadth-first search (BFS) and the latter choice corresponds to depth-first choice (DFS).
+  \label{rmk:graph-exploration-algorithms-termination}
+  Each recursive call in
+  \ref{para:graph-exploration-algorithms-main-idea-3} and
+  \ref{para:graph-exploration-algorithms-main-idea-4}
+  is made on a smaller unexplored pile.
+  We can prove theorems about this algorithm by strong induction on the size of the unexplored pile.
 \end{rmk}
@@
   \label{eg:bfs-example}
-  Consider the following graph:
+  Run BFS on the following graph:
   \begin{center}
@@
         \path (A) edge (C);
-        \path (B) edge (C);
         \path (B) edge (D);
-        \path (B) edge (E);
         \path (C) edge (D);
-        \path (C) edge (E);
         \path (D) edge (E);
@@
   \end{center}
-  We run BFS on this graph.
 
-  (Step 1): The frontier pile is empty and the unexplored pile consists of every vertex in the graph.
-  By \cref{rmk:graph-exploration-algorithms-main-idea-3} we move an arbitrary vertex, e.g, $B$, to the frontier pile.
+  \para
+  Initially, the frontier pile is empty.
+  This matches \ref{para:graph-exploration-algorithms-main-idea-3}, so we move an arbitrary vertex in the unexplored pile to the frontier pile.
   \begin{center}
@@
         \node (A) {$A$};
-        \node[blue] (B) [above right of= A] {$B$};
+        \node (B)[blue] [above right of= A] {$B$};
         \node (C) [below right of= A] {$C$};
@@
 
-  (Step 2): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $B$.
-  We have to move $A$, $C$, $D$, and $E$ to the frontier pile and remove $B$ from the frontier pile.
+  \para
+  The frontier pile is not empty.
+  This matches \ref{para:graph-exploration-algorithms-main-idea-4}.
+  $B$ is adjacent to an explored vertex (e.g., $A$) so we move $A$ to the frontier pile.
   \begin{center}
@@
       \begin{scope}[every node/.style={circle,thick,draw}]
-        \node[blue] (A) {$A$};
-        \node[gray] (B) [above right of= A] {$B$};
-        \node[blue] (C) [below right of= A] {$C$};
-        \node[blue] (D) [right of= B] {$D$};
-        \node[blue] (E) [right of= C] {$E$};
+        \node (A)[blue] {$A$};
+        \node (B)[blue] [above right of= A] {$B$};
+        \node (C) [below right of= A] {$C$};
+        \node (D) [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
         \node (F) [below right of= D] {$F$};
@@
         \path (A) edge (C);
-        \path (B) edge[red] (C);
-        \path (B) edge[red] (D);
-        \path (B) edge[red] (E);
+        \path (B) edge (D);
         \path (C) edge (D);
-        \path (C) edge (E);
         \path (D) edge (E);
@@
 
-  (Step 3): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $A$.
-  We do not have to move anything to the frontier pile because it is not adjacent to any unexplored vertex.
-  Finally, we need to remove $A$ from the frontier pile.
+  \para
+  The frontier pile is not empty.
+  This matches \ref{para:graph-exploration-algorithms-main-idea-4}.
+  $B$ is added to the frontier pile first, so we choose to explore it.
   \begin{center}
@@
       \begin{scope}[every node/.style={circle,thick,draw}]
-        \node[gray] (A) {$A$};
-        \node[gray] (B) [above right of= A] {$B$};
-        \node[blue] (C) [below right of= A] {$C$};
-        \node[blue] (D) [right of= B] {$D$};
-        \node[blue] (E) [right of= C] {$E$};
+        \node (A)[blue] {$A$};
+        \node (B)[blue] [above right of= A] {$B$};
+        \node (C) [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
         \node (F) [below right of= D] {$F$};
@@
         \path (A) edge[red] (B);
-        \path (A) edge[gray] (C);
-        \path (B) edge[red] (C);
+        \path (A) edge (C);
         \path (B) edge[red] (D);
-        \path (B) edge[red] (E);
         \path (C) edge (D);
-        \path (C) edge (E);
         \path (D) edge (E);
@@
 
-  (Step 4): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $C$.
-  Again, we do not have to move anything to the frontier pile because it is not adjacent to any unexplored vertex.
-  Finally, we need to remove $C$ from the frontier pile.
+  \para
+  The frontier pile is not empty.
+  This matches \ref{para:graph-exploration-algorithms-main-idea-4}.
+  Since $B$ is not adjacent to any unexplored vertex, we can move it to the explored pile.
   \begin{center}
@@
       \begin{scope}[every node/.style={circle,thick,draw}]
-        \node[gray] (A) {$A$};
-        \node[gray] (B) [above right of= A] {$B$};
-        \node[gray] (C) [below right of= A] {$C$};
-        \node[blue] (D) [right of= B] {$D$};
-        \node[blue] (E) [right of= C] {$E$};
+        \node (A)[blue] {$A$};
+        \node (B)[gray] [above right of= A] {$B$};
+        \node (C) [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
         \node (F) [below right of= D] {$F$};
@@
         \path (A) edge[red] (B);
-        \path (A) edge[gray] (C);
-        \path (B) edge[red] (C);
+        \path (A) edge (C);
         \path (B) edge[red] (D);
-        \path (B) edge[red] (E);
-        \path (C) edge[gray] (D);
-        \path (C) edge[gray] (E);
+        \path (C) edge (D);
         \path (D) edge (E);
@@
 
-  (Step 5): By \cref{rmk:graph-exploration-algorithms-main-idea-4}, pick the next vertex $D$.
-  We need to move $F$ to the frontier pile and remove $D$ from the frontier pile.
+  %
+
   \begin{center}
@@
       \begin{scope}[every node/.style={circle,thick,draw}]
-        \node[gray] (A) {$A$};
-        \node[gray] (B) [above right of= A] {$B$};
-        \node[gray] (C) [below right of= A] {$C$};
-        \node[gray] (D) [right of= B] {$D$};
-        \node[blue] (E) [right of= C] {$E$};
-        \node[blue] (F) [below right of= D] {$F$};
+        \node (A)[blue] {$A$};
+        \node (B)[gray] [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
       \end{scope}
@@
         \path (A) edge[red] (B);
-        \path (A) edge[gray] (C);
-        \path (B) edge[red] (C);
+        \path (A) edge[red] (C);
         \path (B) edge[red] (D);
-        \path (B) edge[red] (E);
-        \path (C) edge[gray] (D);
-        \path (C) edge[gray] (E);
-        \path (D) edge[gray] (E);
+        \path (C) edge (D);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A)[gray] {$A$};
+        \node (B)[gray] [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (C) edge (D);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A)[gray] {$A$};
+        \node (B)[gray] [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[blue] [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (C) edge (D);
+        \path (D) edge[red] (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A)[gray] {$A$};
+        \node (B)[gray] [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[blue] [right of= C] {$E$};
+        \node (F)[blue] [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[red] (C);
+        \path (B) edge[red] (D);
+        \path (C) edge (D);
+        \path (D) edge[red] (E);
         \path (D) edge[red] (F);
@@
 
-  (Step 6): The unexplored pile is empty.
-  By \cref{rmk:graph-exploration-algorithms-main-idea-2}, we can remove everything from the frontier pile.
+  \para
+  At this point, the unexplored pile is empty, so we have reached the base case \ref{para:graph-exploration-algorithms-main-idea-2}.
   \begin{center}
@@
       \begin{scope}[every node/.style={circle,thick,draw}]
-        \node[gray] (A) {$A$};
-        \node[gray] (B) [above right of= A] {$B$};
-        \node[gray] (C) [below right of= A] {$C$};
-        \node[gray] (D) [right of= B] {$D$};
-        \node[gray] (E) [right of= C] {$E$};
-        \node[gray] (F) [below right of= D] {$F$};
+        \node (A)[gray] {$A$};
+        \node (B)[gray] [above right of= A] {$B$};
+        \node (C)[gray] [below right of= A] {$C$};
+        \node (D)[gray] [right of= B] {$D$};
+        \node (E)[gray] [right of= C] {$E$};
+        \node (F)[gray] [below right of= D] {$F$};
       \end{scope}
@@
         \path (A) edge[red] (B);
-        \path (A) edge[gray] (C);
-        \path (B) edge[red] (C);
+        \path (A) edge[red] (C);
         \path (B) edge[red] (D);
-        \path (B) edge[red] (E);
         \path (C) edge[gray] (D);
-        \path (C) edge[gray] (E);
-        \path (D) edge[gray] (E);
+        \path (D) edge[red] (E);
         \path (D) edge[red] (F);
@@
 
+\begin{eg}
+  \label{eg:dfs-example}
+  Run DFS on the same graph defined in \cref{eg:bfs-example}.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C) [below right of= A] {$C$};
+        \node (D) [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (D);
+        \path (C) edge (D);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  \para
+  This time, let us arbitrarily pick the vertex $C$.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D) [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (D);
+        \path (C) edge (D);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E) [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (D);
+        \path (C) edge[red] (D);
+        \path (D) edge (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  \para
+  Since $D$ is added after $C$, we need to explore $D$ first.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[blue] [right of= C] {$E$};
+        \node (F) [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (D);
+        \path (C) edge[red] (D);
+        \path (D) edge[red] (E);
+        \path (D) edge (F);
+        \path (E) edge (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[blue] [right of= C] {$E$};
+        \node (F)[blue] [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (D);
+        \path (C) edge[red] (D);
+        \path (D) edge[red] (E);
+        \path (D) edge (F);
+        \path (E) edge[red] (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  % 
+  \para
+  At this point, $F$ has been fully explored.
+  We backtrack until we find a vertex that has not been fully explored.
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[blue] [right of= C] {$E$};
+        \node (F)[gray] [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (D);
+        \path (C) edge[red] (D);
+        \path (D) edge[red] (E);
+        \path (D) edge[gray] (F);
+        \path (E) edge[red] (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B) [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[gray] [right of= C] {$E$};
+        \node (F)[gray] [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge (D);
+        \path (C) edge[red] (D);
+        \path (D) edge[red] (E);
+        \path (D) edge[gray] (F);
+        \path (E) edge[red] (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A) {$A$};
+        \node (B)[blue] [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[gray] [right of= C] {$E$};
+        \node (F)[gray] [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge (B);
+        \path (A) edge (C);
+        \path (B) edge[red] (D);
+        \path (C) edge[red] (D);
+        \path (D) edge[red] (E);
+        \path (D) edge[gray] (F);
+        \path (E) edge[red] (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A)[blue] {$A$};
+        \node (B)[blue] [above right of= A] {$B$};
+        \node (C)[blue] [below right of= A] {$C$};
+        \node (D)[blue] [right of= B] {$D$};
+        \node (E)[gray] [right of= C] {$E$};
+        \node (F)[gray] [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge (C);
+        \path (B) edge[red] (D);
+        \path (C) edge[red] (D);
+        \path (D) edge[red] (E);
+        \path (D) edge[gray] (F);
+        \path (E) edge[red] (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+
+  %
+
+  \begin{center}
+    \begin{tikzpicture}[node distance= 2cm]
+      \begin{scope}[every node/.style={circle,thick,draw}]
+        \node (A)[gray] {$A$};
+        \node (B)[gray] [above right of= A] {$B$};
+        \node (C)[gray] [below right of= A] {$C$};
+        \node (D)[gray] [right of= B] {$D$};
+        \node (E)[gray] [right of= C] {$E$};
+        \node (F)[gray] [below right of= D] {$F$};
+      \end{scope}
+      \begin{scope}
+        [every edge/.style={draw= black, very thick},
+        every node/.style={fill= white,circle}]
+        \path (A) edge[red] (B);
+        \path (A) edge[gray] (C);
+        \path (B) edge[red] (D);
+        \path (C) edge[red] (D);
+        \path (D) edge[red] (E);
+        \path (D) edge[gray] (F);
+        \path (E) edge[red] (F);
+      \end{scope}
+    \end{tikzpicture}
+  \end{center}
+\end{eg}
+
 \end{document}

### 2023-11-12 graphs/paper.tex
notation

@@
 
+\begin{notn}
+  The data described in \cref{defn:graphs} can be formalized as a pair that consists of a set of vertices $V$ and a set of edges $E$.
+  We write $(V, E)$ to denote a graph with a vertex set $V$ and an edge set $E$.
+\end{notn}
+
 \begin{eg}
@@
   \label{defn:graphs-undirected}
-  A graph $(V,E)$ is \emph{undirected} if the binary relation $E$ is symmetric.
+  A graph $G = (V,E)$ is \emph{undirected} if the binary relation $E$ is symmetric.
 \end{defn}

### 2023-12-05 dfa.tex
regular lang

\documentclass{zett}
\input{preambles}

\usetikzlibrary {automata, positioning, arrows}
\tikzset{
  ->,
  >=stealth',
  node distance=2cm,
  every state/.style={thick, fill=gray!10},
  initial text=start,
}

\title{regular languages}
\author{Frank Tsai}
\date{\today}
% \thanks{}

\newcommand{\Lfin}{\ensuremath{L_{\mathsf{Fin}}}}

\begin{document}
\maketitle
\tableofcontents

\section{Deterministic finite automaton}
\label{sec:dfa}

\begin{defn}
  A \emph{deterministic finite automaton} (DFA) consists of the following data
  \begin{enumerate}
  \item a nonempty finite set $Q$ of \emph{states};
  \item a nonempty finite set $\Sigma$ of \emph{alphabet};
  \item a \emph{transition function} $\delta : Q \times \Sigma \to Q$;
  \item an \emph{initial state} $q_{0} \in Q$;
  \item a set $F \subseteq Q$ of \emph{final states}.
  \end{enumerate}
\end{defn}

\begin{node}%
  A DFA is a simple \emph{abstract machine}.
  The transition function $\delta : Q \times \Sigma \to Q$ specifies how the machine steps given an input and the current state.
\end{node}

\begin{node}%
  Given a DFA $D = (Q, \Sigma, \delta, q_{0}, F)$.
  We define a new function $\delta^{*} : \Sigma^{*} \to Q$ by recursion:
  \begin{enumerate}
  \item If the input is the empty string then $\delta^{*}(\varepsilon) = q_{0}$;
  \item If the input is a string $s'$ followed by a character $c$ then $\delta^{*}(s' \cdot c) = \delta(\delta^{*}(s'), c)$.
  \end{enumerate}
  For any string $s$ over the alphabet $\Sigma$, $\delta^{*}(s)$ is the state of $D$ after processing $s$.
\end{node}

\begin{defn}
  Let $D = (Q, \Sigma, \delta, q_{0}, F)$ be a DFA.
  We say that a string $s$ over the alphabet $\Sigma$ is \emph{accepted} by $D$ if
  $$
  \delta^{*}(s) \in F
  $$
  In other words, a string $s$ is accepted by $D$ if $D$ ends up in a final state after processing $s$.
\end{defn}

\begin{defn}
  Let $D = (Q, \Sigma, \delta, q_{0}, F)$ be a DFA.
  The \emph{language} of $D$, denoted $\sL(D)$ is the set
  $$
  \sL(D) = \Set{s \in \Sigma^{*} \mid \delta^{*}(s) \in F}
  $$
  We say that a language $L \subseteq \Sigma^{*}$ is \emph{recognized} by a DFA $D$ if $\sL(D) = L$.
\end{defn}

\begin{defn}
  A language $L$ is said to be \emph{regular} if there is a DFA $D$ that recognizes $L$.
\end{defn}

\begin{node}%
  In the following examples, take $\Sigma = \Set{0,1}$.
\end{node}

\begin{eg}
  The empty language $\varnothing$ is regular.
\end{eg}
\begin{node}
  This language is recognized by the following DFA:
  \begin{center}
    \begin{tikzpicture}
      \node[state, initial] (q0) {$q_{0}$};
      \draw (q0) edge[loop above] node{0,1} (q0);
    \end{tikzpicture}
  \end{center}
\end{node}

\begin{eg}
  The language $\Sigma^{*}$ is regular.
\end{eg}
\begin{node}
  This language is recognized by the following DFA:
  \begin{center}
    \begin{tikzpicture}
      \node[state, initial, accepting] (q0) {$q_{0}$};
      \draw (q0) edge[loop above] node{0,1} (q0);
    \end{tikzpicture}
  \end{center}
\end{node}

\begin{eg}\label{eg:finite-languages-regular}
  Every finite language $\Lfin$ is regular.
\end{eg}
\begin{node}
  The idea is that we can construct a DFA by brute force.
  This works because there are only finitely many strings in $\Lfin$, i.e., finitely many states suffice.
  For instance, suppose that $\Lfin = \Set{000, 010, 01}$.
  The following DFA recognizes $\Lfin$:
  \begin{center}
    \begin{tikzpicture}
      \node[state, initial] (q0) {$\lambda$};
      \node[state, right of=q0] (0) {0};
      \node[state, right of=0] (00) {00};
      \node[state, accepting, below of=0] (01) {01};
      \node[state, accepting, right of=01] (010) {010};
      \node[state, accepting, right of=00] (000) {000};
      \node[state, right of=010] (dead) {};
      \draw (q0) edge node[above] {0} (0);
      \draw (q0) edge[out=-45, in=145] node[above, pos=0.1] {1} (dead);
      \draw (0) edge node[above] {0} (00);
      \draw (0) edge node[right, pos=0.3] {1} (01);
      \draw (00) edge node[above] {0} (000);
      \draw (00) edge node[above, pos=0.3] {1} (dead);
      \draw (000) edge node[right] {0,1} (dead);
      \draw (01) edge node[below] {0} (010);
      \draw (01) edge[out=-30, in=-140] node[below, pos=0.2] {1} (dead);
      \draw (010) edge node[below] {0,1} (dead);
      \draw (dead) edge[loop right] node[right] {0,1} (dead);
    \end{tikzpicture}
  \end{center}
\end{node}

\begin{defn}
  Let $L, L' \subseteq \Sigma^{*}$ be languages.
  The concatenation $L \cdot L'$ is the language
  $$
  L \cdot L' = \Set{s \cdot s' \in \Sigma^{*} \mid s \in L \wedge s' \in L'}
  $$
\end{defn}

\begin{thm}\label{thm:regular-closure}
  The class of regular languages is closed under intersection, union, complement, concatenation, and Kleene closure.
  That is, if $L$ and $L'$ are regular languages then so are $L \cap L'$, $L \cup L'$, $\Sigma^{*} \setminus L$, $L \cdot L'$, and $L^{*}$.
\end{thm}
\begin{proof}[Proof sketch]
  \begin{node}\label{node:tensor-of-dfa}%
    We only sketch the proof of closure under union.
    Let $D = (Q, \Sigma, \delta, q_{0}, F)$ (resp., $D' = (Q', \Sigma, \delta', q_{0}', F')$) be a DFA that recognizes $L$ (resp., $L'$).
    The main idea is to construct a new DFA that simulates both $D$ and $D'$:
    \begin{enumerate}
    \item take $Q \times Q'$ as the set of states;
    \item take $\Sigma$ as the alphabet;
    \item define the transition function $\Delta : (Q \times Q') \times \Sigma \to Q \times Q'$ by
      $$
      \Delta((q, q'), c) = (\delta(q, c), \delta'(q', c))
      $$
    \item take $(q_{0}, q_{0}') \in Q \times Q'$ as the initial state;
    \item take $\Set{(q, q') \in Q \times Q' \mid q \in F \vee q' \in F'}$ as the set of final states.
    \end{enumerate}
  \end{node}
\end{proof}

\section{Regular expressions}
\label{sec:regex}

\begin{defn}
  Given an alphabet $\Sigma$, \emph{regular expressions} are defined inductively as follows:
  \begin{enumerate}
  \item $\varnothing$ is a regular expression;
  \item $\varepsilon$ is a regular expression;
  \item each literal $c \in \Sigma$ is a regular expression;
  \item if $r$ and $r'$ are regular expressions then $r \cdot r'$ is a regular expression;
  \item if $r$ and $r'$ are regular expressions then $r + r'$ is a regular expression;
  \item if $r$ is a regular expression then $r^{*}$ is a regular expression.
  \end{enumerate}
\end{defn}

\begin{node}
  We have just defined the syntax of regular expressions.
  The semantics of regular expressions is defined by mapping each regular expression to a language.
\end{node}

\begin{defn}
  The language of a regular expression $r$ is defined by recursion on $r$:
  \begin{enumerate}
  \item the language of the regular expression $\varnothing$ is the empty language:
    $$\sL(\varnothing) = \varnothing$$
  \item the language of the regular expression $\lambda$ is the language containing just the empty string:
    $$\sL(\varepsilon) = \Set{\varepsilon}$$
  \item the language of the regular expression $c$ is the language containing just $c$: for each $c \in \Sigma$,
    $$\sL(c) = \Set{c}$$
  \item the language of the regular expression $rr'$ is the concatenation of $\sL(r)$ and $\sL(r')$:
    $$\sL(r \cdot r') = \sL(r) \cdot \sL(r')$$
  \item the language of the regular expression $r + r'$ is the union of $\sL(r)$ and $\sL(r')$:
    $$\sL(r + r') = \sL(r) \cup \sL(r')$$
  \item the language of the regular expression $r^{*}$ is the Kleene closure of $\sL(r)$:
    $$\sL(r^{*}) = (\sL(r))^{*}$$
  \end{enumerate}
\end{defn}

\begin{thm}\label{thm:regex-to-dfa}
  Let $e$ be a regular expression.
  $\sL(e)$ is regular, i.e., there is a DFA $D$ that recognizes $\sL(e)$.
\end{thm}
\begin{proof}
  \begin{node}
    We proceed by induction on the structure of $e$.
    When $e$ is $\varnothing, \varepsilon,$ or $c$, $\sL(e)$ is finite.
    In \cref{eg:finite-languages-regular}, we informally argued that finite languages are regular.
  \end{node}
  \begin{node}
    When $e$ is $r \cdot r'$ or $r + r'$, the induction hypothesis states that $\sL(r)$ and $\sL(r')$ are regular.
    By definition $\sL(r \cdot r') = \sL(r) \cdot \sL(r')$ and $\sL(r + r') = \sL(r) \cup \sL(r')$.
    By \cref{thm:regular-closure}, the class of regular languages is closed under concatenation and union, so $\sL(r \cdot r')$ and $\sL(r + r')$ are regular.
  \end{node}
  \begin{node}
    Finally, when $e$ is $r^{*}$, the induction hypothesis again states that $\sL(r)$ is regular.
    By definition $\sL(r^{*}) = (\sL(r))^{*}$.
    By \cref{thm:regular-closure}, the class of regular languages is closed under Kleene closure, so $\sL(r^{*})$ is regular.
  \end{node}
\end{proof}

\begin{node}
  \cref{thm:regex-to-dfa} shows that every regular expression has a corresponding DFA.
  The converse is also true: every DFA has a corresponding regular expression.
  It's proof is less trivial.
\end{node}

\end{document}

### 2023-12-05 dfa.tex
edit

@@
     $$\sL(\varnothing) = \varnothing$$
-  \item the language of the regular expression $\lambda$ is the language containing just the empty string:
+  \item the language of the regular expression $\varepsilon$ is the language containing just the empty string:
     $$\sL(\varepsilon) = \Set{\varepsilon}$$
@@
     Finally, when $e$ is $r^{*}$, the induction hypothesis again states that $\sL(r)$ is regular.
-    By definition $\sL(r^{*}) = (\sL(r))^{*}$.
+    By definition, $\sL(r^{*}) = (\sL(r))^{*}$.
     By \cref{thm:regular-closure}, the class of regular languages is closed under Kleene closure, so $\sL(r^{*})$ is regular.
@@
   The converse is also true: every DFA has a corresponding regular expression.
-  It's proof is less trivial.
+  The proof of this fact is not trivial.
 \end{node}
